---
title: "Workshop on microbiome data package in Bioconductor"
abstract: >
  Bioconductor provides significant resources for microbiome data acquisition, analysis, and
  visualization. This workshop introduces `curatedMetagenomicData`, a resource providing uniformly 
  processed taxonomic and metabolic functional profiles for more than 5,000 whole metagenome shotgun 
  sequencing samples from 26 publicly available studies, including the Human Microbiome Project, along with curated participant data. It demonstrates analysis of these data using the `dada2`, `phyloseq`, `treeDA`, `adaptiveGPCA` and `metagenomeSeq` packages for denoising, estimating differential abundance, alpha and beta diversities, ordinations, and other aspects of microbiome data analysis. We also show the `treelapse` and `metavizr` packages for browsing and interactive visualization of microbiome profiles. Together, these packages provide easily linked components for data acquisition and flexible analysis of 16S rRNA and whole metagenome shotgun microbiome profiles. At the end of this workshop, users will be able to access publicly available metagenomic data and to perform differential abundance tests, visualization and
    multivariate statistical analyses of these and other data in R/Bioconductor.
output:
  BiocStyle::html_document:
    number_sections: no
    toc: yes
    toc_depth: 4
vignette: >
    %\VignetteIndexEntry{MicrobiomeWorkshopI}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
isdevel <- Biobase::package.version("Biobase") > 2.37
```


# Install packages

Installing all necessary packages can be accomplished by the following command:

```{r, eval=FALSE}
library(BiocInstaller)
biocLite(c("waldronlab/MicrobiomeWorkshop", "epiviz/metavizr@bioc-workshop"),
         dependencies=TRUE)
```

Double-check that "metavizr" is installed from GitHub:
```{r NeededPackagespac}
if(!("metavizr" %in% installed.packages()) || "1.1.2" != installed.packages()["metavizr","Version"])
  devtools::install_github("epiviz/metavizr@bioc-workshop", build_vignettes = TRUE)
```

Load all needed packages:
```{r}
cranpkgs=c("ggplot2","devtools","vegan","httr")
BioCpkgs=c( "curatedMetagenomicData", "phyloseq")
ghpkgs= "metavizr"
allpkg <- c(cranpkgs, BioCpkgs, ghpkgs)
sapply(allpkg, require, character.only = TRUE)
```

# `curatedMetagenomicData`

`curatedMetagenomicData` provides 6 types of data for each dataset:

1. Species-level taxonomic profiles, expressed as relative abundance from kingdom to strain level
2. Presence of unique, clade-specific markers
3. Abundance of unique, clade-specific markers
4. Abundance of gene families
5. Metabolic pathway coverage
6. Metabolic pathway abundance

Types 1-3 are generated by 
[MetaPhlAn2](http://huttenhower.sph.harvard.edu/metaphlan2); 4-6 are generated by [HUMAnN2](http://huttenhower.sph.harvard.edu/humann2).

Currently, `curatedMetagenomicData` provides:

* 5,716 samples from 26 datasets, primarily of the human gut but including body sites profiled in the Human Microbiome Project
* Processed data from whole-metagenome shotgun metagenomics, with manually-curated metadata, as integrated and documented Bioconductor 
ExpressionSet objects
* ~80 fields of specimen metadata from original papers, supplementary files, and websites, with manual curation to standardize annotations
* Processing of data through the [MetaPhlAn2](http://huttenhower.sph.harvard.edu/metaphlan2) pipeline for taxonomic abundance, and [HUMAnN2](http://huttenhower.sph.harvard.edu/humann2) pipeline for metabolic analysis
* This effort required analyzing ~100T of raw sequencing data

These datasets are documented in the reference manuals of the [release](http://bioconductor.org/packages/release/data/experiment/html/curatedMetagenomicData.html) or [devel](http://bioconductor.org/packages/devel/data/experiment/html/curatedMetagenomicData.html) versions of Bioconductor. The development version has approximately twice as many datasets and significantly more functionality. The AMI provided at the Bioc2017 workshop runs the development version of Bioconductor; see http://bioconductor.org/developers/how-to/useDevel/ for instructions on upgrading your own machine to the development version and switching between devel and release.

## Using `curatedMetagenomicData` Resources

Use of the resources in `curatedMetagenomicData` is simplified with the use of Bioconductor's ExperimentHub platform, which allows for the accessing of data through an intuitive interface. First, `curatedMetagenomicData` is installed using *BiocInstaller* and then called as a library - the process allows for the user to simply call datasets as functions because the package is aware of the resources present in ExperimentHub S3 buckets. 

```{r, message=FALSE}
# BiocInstaller::biocLite("curatedMetagenomicData")  #Bioconductor version
# BiocInstaller::biocLite("waldronlab/curatedMetagenomicData")  #bleeding edge version
suppressPackageStartupMessages(library(curatedMetagenomicData))
```

## Available samples and metadata

The manually curated metadata for all available samples are provided in a single table `combined_metadata`:

```{r eval=FALSE}
?combined_metadata
View(combined_metadata)
```

```{r}
table(combined_metadata$antibiotics_current_use)
table(combined_metadata$disease)
```


## Accessing datasets

Individual data projects can be fetched via per-dataset functions or the `curatedMetagenomicData()` function. A function call to a dataset name returns a Bioconductor ExpressionSet object:

```{r, message=FALSE}
suppressPackageStartupMessages(library(curatedMetagenomicData))
zeller.eset = ZellerG_2014.metaphlan_bugs_list.stool()
```

The following creates a list of two `ExpressionSet` objects providing the BritoIL_2016 and Castro-NallarE_2015 oral cavity taxonomic profiles (this method works in release or devel, but these datasets are available in devel only). See a complete list of available datasets in the PDF manual.

```{r, message=FALSE, results='hide', eval = isdevel}
oral <- c("BritoIL_2016.metaphlan_bugs_list.oralcavity",
          "Castro-NallarE_2015.metaphlan_bugs_list.oralcavity")
esl <- curatedMetagenomicData(oral, dryrun = FALSE)
esl
esl[[1]]
esl[[2]]
```

And the following would provide all stool metaphlan datasets if `dryrun = FALSE` were set:

```{r, eval=TRUE}
curatedMetagenomicData("*metaphlan_bugs_list.stool*", dryrun = TRUE)
```

### Merging multiple datasets

The following merges the two oral cavity datasets downloaded above into a single ExpressionSet (devel only). 

```{r, eval=isdevel}
eset <- mergeData(esl)
eset
```

This works for any number of datasets. The function will not stop you from merging different data types (e.g. metaphlan bugs lists with gene families), but you probably don't want to do that.

## Using `ExpressionSet` Objects

All datasets are represented as `ExpressionSet` objects because of the integrative nature of the class and its ability to bind data and metadata. There are three main functions, from the `Biobase` package, that provide access to experiment-level metadata, subject-level metadata, and the data itself.

To access the experiment-level metadata the `experimentData()` function is used to return a `MIAME` (Minimum Information About a Microarray Experiment) object.

```{r}
experimentData( zeller.eset )
```

To access the subject-level metadata the `pData()` function is used to return a `data.frame` containing subject-level variables of study.

```{r}
head( pData( zeller.eset ) )
```

To access the data itself (in this case relative abundance), the `exprs()` function returns a variables by samples (rows by columns) numeric matrix. Note the presence of "synthetic" clades at all levels of the taxonomy, starting with kingdom, e.g. k__bacteria here:

```{r}
exprs( zeller.eset )[1:6, 1:5]  #first 6 rows and 5 columns
```

Bioconductor provides further documentation of the ExpressionSet class and has published an excellent [introduction](https://tinyurl.com/ExpressionSetIntro).

## Creating `phyloseq` and `metagenomeSeq` objects

`curatedMetagenomicData` provides convenience functions for converting its *ExpressionSet* objects to *phyloseq::phyloseq* and *metagenomeSeq:MRExperiment* class objects for downstream analysis. For example, to convert the **zeller.eset** object to these two classes:

```{r}
ExpressionSet2phyloseq(zeller.eset)
ExpressionSet2MRexperiment(zeller.eset)
```

This also works on the merged **eset** object from above:

```{r, eval=isdevel}
ExpressionSet2phyloseq(eset)
ExpressionSet2MRexperiment(eset)
```

## phylogenetic trees and UniFrac distances

Set **phylogenetictree = TRUE** to include a phylogenetic tree in the `phyloseq` object (bioc-devel only):

```{r}
zeller.tree <- ExpressionSet2phyloseq( zeller.eset, phylogenetictree = TRUE)
```

```{r}
wt = UniFrac(zeller.tree, weighted=TRUE, normalized=FALSE, 
             parallel=FALSE, fast=TRUE)
plot(hclust(wt), main="Weighted UniFrac distances")
```

## Exploratory analysis of E. coli prevalence

Here's a direct, exploratory analysis of *E. coli* prevalence in the zeller dataset using the `ExpressionSet` object. More elegant solutions will be provided later using subsetting methods provided by the `r BiocStyle::Biocpkg("phyloseq")` package, but for users familiar with `grep()` and the `ExpressionSet` object, such manual methods may suffice.

First, which *E. coli*-related taxa are available?

```{r}
grep("coli", rownames(zeller.eset), value=TRUE)
```

Create a vector of *E. coli* relative abundances. This `grep` call with a "$" at the end selects the only row that ends with "s__Escherichia_coli":
```{r}
x = exprs( zeller.eset )[grep("s__Escherichia_coli$", rownames( zeller.eset)), ]
summary( x )
```

This could be plotted as a histogram:
```{r}
hist( x, xlab = "Relative Abundance", main="Prevalence of E. Coli",
      breaks="FD")
```

## Estimating Absolute Raw Count Data

Absolute raw count data can be estimated from the relative count data by multiplying the columns of the `ExpressionSet` data by the number of reads for each sample, as found in the `pData` column "number_reads". For demo purposes you could (but don't have to!) do this manually by dividing by 100 and multiplying by the number of reads:

```{r}
zeller.counts = sweep(exprs( zeller.eset ), 2, zeller.eset$number_reads / 100, "*")
zeller.counts = round(zeller.counts)
zeller.counts[1:6, 1:5]
```

or just set the `counts` argument in `curatedMetagenomicData()` to `TRUE`:

```{r}
zeller.eset2 = curatedMetagenomicData("ZellerG_2014.metaphlan_bugs_list.stool",
                                     counts = TRUE, 
                                     dryrun = FALSE)[[1]]
all.equal(exprs(zeller.eset2), zeller.counts)
```


# Taxonomy-Aware Analysis using `phyloseq`

For the MetaPhlAn2 bugs datasets (but not other data types), you gain a lot of phylogeny-aware, ecological analysis and plotting by conversion to a `r BiocStyle::Biocpkg("phyloseq")` class object. `r BiocStyle::Biocpkg("curatedMetagenomicData")` provides the `ExpressionSet2phyloseq()` function to make this easy:

```{r, warning=FALSE}
suppressPackageStartupMessages(library(phyloseq))
zeller.pseq = ExpressionSet2phyloseq( zeller.eset )
```

# Visualization with Metavizr

We can now examine the zeller dataset with Metavizr. This visualization tool allows interactive exploration of the taxonomic hierarchy and statistically-guided visual analysis. Specifically, metavizr enables feature selection through a navigation mechanism called FacetZoom. First we need to format the MRexperiment object to be used by metavizr. Currently, metavizr expects an annotation for each leaf, so this removes the inner nodes of the hierarchy. A version of metavizr is in development to use the inner node count values output by ExpressionSet2MRExperiment directly.

```{r}
mr_exp_zeller <- ExpressionSet2MRexperiment(zeller.eset)

fd = fData(mr_exp_zeller) 
for(i in seq(ncol(fd))){
   fd[,i] = as.character(fd[,i]) 
} 
fData(mr_exp_zeller) = fd

leaves_data <- mr_exp_zeller[-unname(which(is.na(fData(mr_exp_zeller)$Species))),]

na_strain_indices <- which(is.na(fData(leaves_data)$Strain))

for(n in na_strain_indices){
  fData(leaves_data)[unname(n),]$Strain <- paste("Not_Annotated_Strain", paste(fData(leaves_data)[unname(n),]$Species, sep = ""), sep = "_")
}
```

Once the data is formatted, we can start a WebSocket connection to an instance of the Metaviz frontend. Then we add the navigation utility along with an interactive boxplot.

```{r, eval=FALSE}
feature_order <- colnames(fData(leaves_data))

public_ipv4 <- try(httr::content(httr::GET("http://169.254.169.254/latest/meta-data/public-ipv4")), silent = TRUE)

if(grepl("Timeout was reached", simpleMessage(public_ipv4))){
  app <- startMetaviz()
} else{
  app <- startMetavizStandalone(host=public_ipv4, daemonized=FALSE)
}

app <- startMetavizStandalone(host=public_ipv4, daemonized=FALSE, verbose = TRUE)

facetZoom <- app$plot(leaves_data, datasource_name = "zeller", feature_order = feature_order)
app$service()
heatmap_plot <- app$chart_mgr$revisualize(chart_type = "HeatmapPlot", chart = facetZoom)
app$service()

#On rstudio-server version < 1.1.273, the user need to run the app$service() command after each user input on the UI. There is a bug with Rstudio-server interferring with the R event loop. metavizr should run without issue on rstudio desktop.
```

From the metavizr Vignette: You should now see a FacetZoom to explore the hierarchy of the metagenomic features from the MRexperiment object. To navigate the complex, hierarchical structure of the feature space, we made use of the FacetZoom visualization design and extended it to metagenomic data. Because of the limitations in the screen size and performance rendering big trees, this technique is an efficient visualization for navigating trees. The FacetZoom helps zoom in and out of trees and traverse subtrees. Every node in the tree has a state associated with it and are three possible states for a node 1) expand - use all subtree nodes during analysis 2) collapse - aggregate all nodes under the subtree to the selected node 3) remove - discard all nodes in the subtree for the analysis. The state of a node is also propagated to all its children and can be identified by the opacity of the node. Row level controls are available on the left side - to set the state of all nodes at a selected depth/taxonomic class of the hierarchy. The right hand side of the FacetZoom shows the location of the current subtree. Users can set states on the nodes to define a cut over the feature space. The cut defines how the count data is visualized in linked plots and charts. In addition to defining the cut, we also have a navigation bar to limit the range of features when querying count data. Navigation controls move the bar left/right and extend over the entire range of the current tree in the icicle. Each of these actions queries the underlying data structure and automatically propagates the changes to other visualizations in the workspace. When navigating outside the scope of the navigation bar, chevrons (left/right) appear on the navigation bar to help identify the current position. Hovering over a node highlights the entire lineage in the FacetZoom along with all other linked charts and plots.

#TODO: Feature selection with metagenomeSeq and visual selections in metaviz
